---
title: "Effect of smoking on celiac disease: results from a synthetic case-control population"
subtitle: "Spring 2023 Obligatory Assignment; SMED8002 Epidemiology II; " 
date: 2023-03-20
date-format: long
author: 
  - name: Author1
  - name: Author1
  - name: Author1
  - name: Mohammad Sayeef Alam

toc: true
number-sections: true
logo: ntnu_bredde_eng.png
bibliography: smed8002ref.bib
csl: apa-single-spaced.csl
html-math-method: katex
format: 
  html:
    mainfont: Times New Roman
    fontsize: normal
  pdf: default
---

```{r libs, echo=F}
librarian::shelf(tidyverse, dagitty, ggdag, ggplot2, quiet = T)
```

## Introduction


Celiac disease (CD) has become globally common in all age groups [@vilppula2009increasing], especially among Caucasian populations who are genetically predisposed. While the global prevalence of CD is around 1% [@singh2018global], some Nordic populations have recorded higher than global average numbers [@vilppula2009increasing; @kvamme2022population; @dydensborg2012increasing; @myleus2009celiac]. In fact, only about one in every hundred people from the Western countries eventually develop the disease even though the consumption of gluten is remarkably high.

The prevalence of CD has significantly increased over the past 50 years and in the northern Europe it is prevalent among 1-2% of the population [@lohi2007increasing].

## Methods

A synthetic data has been generated using the HUNT 4 samples as reference. The outcome variable is the occurrence of celiac disease assessed by positive serological or biopsy results. The exposure of interest in the study was the smoking status of the individual. Socio-economic factors such as age and sex were considered with behavioral factors such as alcohol and coffee consumption as confounding between the exposure and outcome.

```{r generate_data, echo=F}
# generate synthetic vectors ----
set.seed(111222)

iid = round(floor(runif(54051)*10000000), 0)
sex = sample(c("Female","Male"), 54051, replace = T, prob = c(0.54, 0.46))
age = rnorm(54051,54.9,17.8)
wgt = rnorm(54051,80.04,16.43)
coffee = sample(c("Never","1 cup/day","2-3 cups/day","4 cups or more/day"), 54051, replace = T, prob = c(0.17, 0.17, 0.38, 0.28))
alcohol = sample(c("No","Yes"), 54051, replace = T, prob = c(0.169, 0.831))
smoking = sample(c("Never", "Ever"), 54051, replace = T, prob = c(0.44, 0.56))
educ = sample(c("Grunnskole","1-2årig videregående skole","3 år i videregående skole","Fagbrev eller svennebrev", "Høyskole/universitet, mindre enn 4 år","Høyskole/universitet, 4 år eller mer"), 54051, replace=TRUE, prob=c(0.117, 0.156, 0.123, 0.214, 0.204, 0.185))
hhincome = sample(c("Under 250 000 kr", "250 000 – 450 000 kr", "451 000 – 750 000 kr", "751 000 – 1 000 000 kr", "Over 1 000 000 kr"), 54051, replace=TRUE, prob=c(0.103, 0.216, 0.29, 0.212, 0.176))
celiac = sample(c("No","Yes"), 54051, replace = T, prob = c(0.982, 0.018))

# cbind the vectors into dataframe ----
syndata = data.frame(iid,sex,age,wgt,coffee,alcohol,educ,hhincome,smoking,celiac)

# data manipulation ----
# convert the categorical variables into factors
syndata = syndata %>% mutate_at(c("sex", "coffee", "alcohol", "smoking", "hhincome", "educ", "hhincome", "celiac"), factor)

# write the data into a csv for future usage
write.csv(syndata, "C:/Users/mohasal/OneDrive - NTNU/Courses/SMED8002/syndata.csv", row.names = F)
```

## Results

```{r ac1, echo=F}
# analysis ----
# import the data
cscrtldata = read.csv("C:/Users/mohasal/OneDrive - NTNU/Courses/SMED8002/syndata.csv")
```

```{r, echo=FALSE}
#| label: dag
#| fig-cap: "Figure 1: Directed Acyclic Graph"

# make a DAG diagram
dag_diag = dagify(celiac ~ smoking + age + sex + alcohol + coffee +  wgt + educ,
  smoking ~ age + sex + alcohol + coffee +  wgt + educ,
  exposure = "smoking",
  outcome = "celiac",
  labels = c(
    "coffee" = "Coffee consumption",
    "smoking" = "Smoking",
    "age" = "Age",
    "sex" = "Sex",
    "wgt" = "Weight",
    "alcohol" = "Alcohol consumption",
    "celiac" = "Celiac Disease",
    "educ" = "Highest  education"
  )
)

ggdag_parents(dag_diag, "smoking", text = FALSE, use_labels = "label") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none"
        )
```

To refer @dag do as the code suggest.

## Discussion

There are sufficient evidence that environmental exposures increases the risk of celiac disease.
