---
title: Statisitcal testing vs Estimation
subtitle: "Chapter 8: Random Error and the Role of Statistics"
description: HUNT Forskningslunsj
footer: https://github.com/alamsayeef/quarto/tree/main/presentations
author: 
  - name: Mohammad Sayeef Alam
    affliation:
      - name: K. G. Jebsen Center for Genetic Epidemiology and HUNT Research Center
        address: NTNU, Trondheim
date: 2023-03-31
date-format: long
toc-title: Contents
toc: true
format: 
  revealjs:
    chalkboard: true
    logo: HUNTlogo.png
    theme: serif
    slide-number: c/t
    incremental: true
---

## p-value functions

*   A.k.a. Confidence interval function
*   p-value is compatibility between data and H0
*   RR lies [0,$\infty$]
*   If outcome is negative: 
    +   RR < 1 indicates lower risk
    +   RR > 1 indicates higher risk
*   Hypothesis testing on multiple values
*   p-values corresponding for each RR generates a **p-value function**


:::{.notes}
-   aka confidence interval function
-   The p-value is a statistic that can be viewed as a measure of the compatibility between the data in hand and the null hypothesis.
-   Any rate ratio measure is taken which ranges between 0 - infinity, our H0 is RR = 1, i.e. no effect of exposure on outcome. (If it would have been a negative outcome, RR < 1 would indicate protective effect/lower risk and RR > 1 indicates higher risk)
-   What if we could test for multiple H0 hypotheses
-   calculate an infinite number of p-values that test every possible value of RR leads to generation of a p-value function
:::
	
## Example 1

Case-control study of drug exposure during pregnancy and congenital heart disease.

```{r lib, echo=F, include=F, warning=F}
librarian::shelf(fmsb, pvaluefunctions)
```
:::: {.columns}

::: {.column width="40%"}
```{r df1, echo=F}
df1 = as.data.frame(matrix(c(4, 386, 4, 1250), 2, 2), row.names = c("Cases", "Controls"))
print(df1)
```
:::

::: {.column width="60%"}
```{r cc1, echo=F}
p1 = pvalueplot(matrix(c(4, 386, 4, 1250), 2, 2), xrange=c(0.1, 30), plot.OR = T)
```
:::

::::

:::{.notes}
-   Resembles a Tepee [image]
-   Table & figure
-   p-value = 1 is the location where the data is most compatible with the estimate; RR = 3.2
-   The curve provides a quantitative overview of the statistical relation between exposure and disease.
-   degree of precision of the estimate, which is indicated by the narrowness or the breadth of the tepee.
-   when we see the p-value for RR = 1 we observe it to be 0.08, meaning there is no statistically significant association between exposure and outcome.
-   Hold up! what? this is very confusing when understanding it in a reverse fashion.
-   What if our p-value was 0.02, we would have rejected the H0 and said there is statistical significance between the effect of exposure and outcome.
-     In other words,
		    when p-value < 0.05: Reject H0 and accept H1
		    when p-value > 0.05: Reject H1 and accept H0
-   The p-value for RR = 1 is identical to the p-value for RR = 10.5, so
there is no reason to prefer the interpretation of RR = 1 over the interpretation
that RR = 10.5.
-   Better than both is the choice of RR = 3.2
-   Take away is that p-value can be gravely misleading !sometimes!
:::

## Example 2

Hypothetical case-control data

:::: {.columns}

::: {.column width="40%"}
```{r df2, echo=F}
df2 = as.data.frame(matrix(c(1090, 14910, 1000, 15000), 2, 2), col.names= c("With exposure", "Without exposure"), row.names = c("Cases", "Controls"))
print(df2)
```
:::

::: {.column width="60%"}
```{r cc2, echo=F, warning=F}
p2 = pvalueplot(matrix(c(1090, 14910, 1000, 15000), 2, 2), xrange=c(0.1, 2), plot.OR = T)
```
:::

::::


:::{.notes}
-   The graph clearly indicates that there is no strong effect btw exposure and outcome
-   Yet due to p-value being 0.04 we reject the H0: RR = 1 and say that there is statistical evidence of a significant effect between the exposure and outcome.
-   Interpretation of data purely based on significance testing would be indicative of one's perverse behaviour.
-   p-value function provides a quantitative visual message about the size of the effect.
-   Strength = Location; Precision = 1/Spread
-   One number does not explain two quantities
:::

## CI

:::{.notes}
-   The LCL and UCL act as the indicators of precision
-   Very uncommon to see a full p-valu function/plot but pretty common to have CI in epidemiological studies
-   Depending upon your tolerance you can choose the CI to be between 80%, 90%, 95% or even 99%
-   Surrogate for statistical significance; rather should be viewed as counterpart of p-value function
-   Should be interpreted as how well the data and estimate agree with each other.
:::

## Problems

Example 1

---

Example 2

---

Example 3

## Formulas for calculations

-   Incident Rate Difference (RD)
-   CI % | z value
    
    90	  1.645
    
    95	  1.96
    
    99	  2.57

-   Considering a 90% CI
		RDl,RDu = RD +- 1.645*SD; applicable to any measure of difference
-   When it comes to ratios, the measures usually tend to be assymetrically towards larger values; to counter it, we log transform.
		ln(RDl),ln(RDu) = ln(RD) +- 1.645*ln(SD)
-   Since these are in logarithmic scale after calculation we need to revert back to original values by exponentiating it with the base of nature logarithm
		RDl,RDu = e^(ln(RD) +- 1.645*ln(SD))

:::{.notes}

:::

## Take home message

-   p-value superfical, sophisticated application of quantitative method with only a qualitative proposition
-   "Significant" or "Not significant" provides no quantitative clue about the size of the effect
-   Lack of significance DOES NOT imply lack of effect or Significance DOES NOT imply "proof" of effect either
-   Non chance causation alt theories: confounding, non differential classification (selection/measurement bias).
-   When one arrives at insignificant results one should definitely probe deeper for associations missed by chance due to reasons mentioned above
-   Confidence limits should be gray areas used to mentally picture the p-value function and hence the agreement btw data and estimate.

## Tusen takk {background-color="#447099" .unlisted}

Håper å sende deg hjem med mindre forvirring enn du kom i morgen :) 

Beklager hvis det er noe annet :(